Composition-climate paper

### 2016-06-17 Composition_Ecology.Rmd
Finished key text edits.  Will go through a printed version to check over tomorrow morning.  Added references, still missing full citation to @IceAgeEcologist & Burke paper.  Removed duplicate reference.
Added Early & Sax referecne & mention early in Introduction.

Increasingly, ecological models and their predictive capacity are being improved by integrating contemporary observational data with paleohistorical data, when available [@moritz2013future;@maguire2015modeling].  Future projections that rely only on contemporary observational data are limited by a narrow time window that may obscure greater plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches that lie outside the current realized environmental space [@jackson2000responses], miss the responses of species to past no-analog climates [@veloz2012no], and fail to detect the effects of historical legacies and contingency on species distributions [@svenning2015influence]. Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence]. Habitat loss through human agency can impact apparent species distributions at timescales of thousands [@sarmento2013impacts] to hundreds of thousands of years [@ivory2016niche]. In principle, dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lage climate change.  In practice, however, parameterization of these models is critical, and processes operating at large temporal and spatial scales remain poorly constrained [@moorcroft2006close].

@@

`r fig_nums("fourPanel", "(a) Diagram of the 2x2 factorial design used to diagnose the effects of historic climate and vegetation change on the observed shifts in tree-climate relationships. We analyzed observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row). (b-e) Representative normalized probability densities for Larix for each of the four combinations of climate and vegetation states, plotted relative to Tmax (b), Tdif (c), Tmin (d), and Pann (e).")`

@@

`r fig_nums("gain_loss", "Maps showing areas of gain, loss, and continuous presence of tree genera. 'Loss' indicates 8x8km grid cells where PLSS data indicated presence of a genus but FIA data do not record presence (light blue). 'Gain' indicates areas where a genus is absent in the PLSS data but present in the FIA data (red). 'Continuous presence' represents locations where both FIA and PLSS data indicate presence (dark gray). Losses are more common than gains, particularly in the southwestern portion of tree ranges.  Few taxa show gains, although Populus, Fraxinus and Tilia show gains of over 10% (Table 2).")`

@@

For most tree genera, historic range losses dominate over range gains and are spatially structured (Figure 3). Range losses in the Upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3). For Ulmus and Larix, however, losses occur throughout the range.  Losses are highest for Ostrya/Carpinus and Larix when assessed as a proportion of total range (Table 2), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.  Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed for Populus (Table 2) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.   Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).  Range gains for *Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in a relatively few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plots and stand heterogeneity, rather than systematic range gains.  Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.  *Pinus*, *Thuja*, and *Ulmus* gains are limited but primarily in the southern portion of their historic range, or beyond it.

@@

Both the total change in realized climate distributions and the attribution of these changes to historic land use or climate change varies strongly among taxa and among climate variables (Figure 5). Shifts in realized climate distributions over the last 100 to 150 years are largest for Ulmus (TJan, TJul), Larix (Tmax), Tsuga (Tmax and Pann), and Fagus (Pann).   Attribution analyses based on the d index (Figure 5) reveal that while many genera show the imprint of historic land use on shifting climatic niches, this shift is not uniform across taxa. Maximum temperature shows the greatest land use signal (Figure 5X), with effects particularly strong for Larix, Ulmus, and Fagus. The large shift in the distribution of Tsuga relative to July temperatures, conversely, appears to be mainly attributable to historic climate change. For minimum temperature (Figure 5X), shifting distributions of Ulmus and Cedar can be mainly attributed to land use change, while Larix and Betula have large shifts that are equally attributable to historic climate change and land use, and the niche shifts for other taxa are driven by changing climate.  The climatic signal is strongest for annual precipitation, with regional changes in annual precipitation being largely responsible for shifting climate niches (Figure 5X), while only Populus and Picea have seen climate niche shifts due to land-use that are larger than the climate shifts.

@@

This work demonstrates that historic land use can substantially alter the realized climate niches of species and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat]. Land use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000global]. While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@goring2015preset], and the correlational structure between species and climate. Paleoclimatic modeling indicates shifts in realized niches for species over time, potentially resulting from non-analogue climates or shifts in competition resulting from changing patterns of co-occurence through time [@maiorano2013building].

Shifts in realized climate space since Settlement may have important implications for the management of forest communities, and the fact that historical realized niches are broader than modern niches might lead us to be more optimistic than otherwise, however, projected climate change far exceeds the shifts due to historical land use and climate change.  In addition, coupled migration and adaptation in response to future climate change is likely to be impeeded by the exitrpation of communities due to land use change.  Species migration and genetic adaptation are coupled across time-scales [@davis2001range].  The loss of forest communities along the forest-prairie boundary in the Upper Midwest is particularly important given that the direction of climate change regionally is towards warmer conditions, conditions that exists at the forest-prairie boundary.  The loss of genotypes adapted to warmer and drier conditions (the lagging edge of distributions for many taxa in the Upper Midwest) may mean more rapid losses at the lagging edges of the species range.  Long distance gene flow of adaptive traits to central and leading populations from the lagging edge [@hu2006seed] is lost when land use conversion to agricultural production results in the extirpation of locally adapted populations.  Thus net gene flow in populations along the remaining "lagging" edge will be from the central and leading populations, individuals maladapted to warmer conditions. This maladaptive gene flow could potentially amplify the effects of climate change on populations along the forest-prairie boundary leading to increasing risk of local extinction among natural populations [@kremer2012long].

From a modern perspective, the US Forest Service's Forest Inventory and Analysis data set is an important resource for researchers, providing detailed data at fine grain over the entire contiguous United States. However, the FIA is also an incomplete record of tree species distributions in the United States since it limits plot surveys to plots with > 10% tree cover [@woudenberg2010forest].  In much the same way, climatic data plays an important role in building correlative models, but we are limited to the modern era [@pellatt2012using;@iverson1998predicting], while evidence strongly suggests historical datasets add considerably to our understanding of the breadth of species' climatic niches [@ivory2015; @maiorano2013building;@early2014climatic].

The use of correlational species-distribution models may be heavily impacted by historical land-use change and the sampling design of modern survey data.  Dorman et al. [@dormann2012correlation] propose a gradient of models, from correlative to process-based models for predicting species distributions.  While process based models may provide more accurate predictions of species distributions given the loss of spatial coverage as a result of land-use conversion, spatially biased land-use conversion could result in the extirpation of well-adapted genotypes within a population.  The inability to acurately reflect phenotypic variability along gradients of adaptation could then lead to weaker process-based models.  In this case, we might assume that both correlative and process based models would be affected by spatially biased land-use conversion, particularly at "lagging" margins of a species distribution.

At a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations that might otherwise not support forest cover [@brazel2000tale]. Human agency in these marginal environments ultimately leads urban ecosystems in deserts to closely resemble those in the northeastern United States [@groffman2014ecological], but may also support the extension of forest ecosystems at historical range margins, or preserve historical range margins from land use conversion to agriculture as windrows, or isolated woodlots.  Detailled sampling of tree species, particularly in urban and agricultural environments beyond the current expected range limits of species may provide broader coverage of a species' fundamental climate niche, but care must be taken to account for the intensive management that may be required to maintain species beyond their realized niche.

This work reinforces the importance of historical datasets for examining the relationships between species and climate, and their importance in forecasting species distributions in the future.  We show that modern inventory data samples only a subset of historical distribution, both spatially and within multidimentsional climate space.  Critically, we show that the combination of regional land-use change and climate change leads to the under-prediction of species resilliance to climate change regionally.  Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 from a 1980s baseline [@wicci2011climate].  Every tree genera examined in this study had historical $T_{max}$ values between 0.4 and 1.3^o^C higher than modern estimates, and with $P_{ann}$ between 42 and 86mm lower than modern estimates.  Projected changes in $P_{ann}$ range from -17 - 110mm from a 1980 baseline to 2055 based on fourteen global climate models compiled as part of the CMIP3 intercomparison [@wicci2011climate]. Higher temperatures and lower precipitation means for taxa in the historic period indicate the potential for more robust responses to climate change than might be assessed using modern data only.

Increasing forest fragmentation as a result of land use conversion overlies long term climate patterns.  In regions where land use conversion is climatically biased, conversion can affect the realized niche of tree species, particularly when sampling of those species is biased by a survey design that is itself biased by land use converison.  In the Upper Midwestern United States maximum temperatures have declined over the last century, but increases in minimum annual temperature, and land use conversion along the warm and dry forest-prairie margin have resulted in shifts that compound the apparent climatic shifts (`r tab_nums("compTable", display = "cite")`), resulting in a realized niche for most species that is less robust to future shifts in climate, toward warmer conditions with, potentially higher evapotranspiration.

While historical datasets may add to our ability to understand species realized climate niches and their adaptive potential in the face of climate change they also serve to underscore the important interactions between climate change and land use change at a population level.  The homogenization of midwestern forests evidenced through forest composition data [@goring2015preset;@rhemtulla2009legacies;@schulte2007homogenization;@hanberry2012comparison] may well be reflected in genetic data, although this is less well understood, and may be lost to us due to the broad-scale disturbance resulting from EuroAmerican Settlement in the region.  Ultimately, land-use history is a critical component of modern ecology that is under-represented in the literature [although see: @hamburg1986disturbance;@compton2000long;@foster1998land].

>>>>

Increasingly, ecological models and their predictive capacity are being improved by integrating contemporary observational data with paleohistorical data, when available [@moritz2013future;@maguire2015modeling].  Future projections that rely only on contemporary observational data are limited by a narrow time window that may obscure greater plasticity in species-climate responses [@maiorano2013building], incompletely delimit species fundamental niches that lie outside the current realized environmental space [@jackson2000responses], miss the responses of species to past no-analog climates [@veloz2012no], and fail to detect the effects of historical legacies and contingency on species distributions [@svenning2015influence]. Paleoecological records from the last deglaciation show that the time lag of forest responses to abrupt climate change varies from a few years to centuries [@williamsinpress;@webb1986vegetation;@ordonez2013realized;@svenning2015influence]. Habitat loss through human agency can impact apparent species distributions at timescales of decades [@early2014climatic], thousands [@sarmento2013impacts] to hundreds of thousands of years [@ivory2016niche]. In principle, dynamic vegetation models [*e.g.*, @leiblein2016modelling] can represent the physiological, demographic, and dispersal processes that cause tree populations to lage climate change.  In practice, however, parameterization of these models is critical, and processes operating at large temporal and spatial scales remain poorly constrained [@moorcroft2006close].

@@

`r fig_nums("fourPanel", "(a) Diagram of the 2x2 factorial design used to diagnose the effects of historic climate and vegetation change on the observed shifts in tree-climate relationships. We analyzed observed and hypothetical tree-climate relationships for one of four combinations of pre-settlement vegetation (pink fill, left column) or modern vegetation (blue fill, right column) and pre-settlement climate (solid border, top row) or modern climate (dashed border, lower row). (b-e) Representative normalized probability densities for *Larix* for each of the four combinations of climate and vegetation states, plotted relative to Tmax (b), Tdif (c), Tmin (d), and Pann (e).")`

@@

`r fig_nums("gain_loss", "Maps showing areas of gain, loss, and continuous presence of tree genera. 'Loss' indicates 8x8km grid cells where PLSS data indicated presence of a genus but FIA data do not record presence (light blue). 'Gain' indicates areas where a genus is absent in the PLSS data but present in the FIA data (red). 'Continuous presence' represents locations where both FIA and PLSS data indicate presence (dark gray). Losses are more common than gains, particularly in the southwestern portion of tree ranges.  Few taxa show gains, although *Populus*, Fraxinus and Tilia show gains of over 10% (Table 2).")`

@@

For most tree genera, historic range losses dominate over range gains and are spatially structured (Figure 3). Range losses in the Upper Midwest are concentrated in the south and west, along the prairie-savanna-forest ecotone (Figure 3). For *Ulmus* and *Larix*, however, losses occur throughout the range.  Losses are highest for Ostrya/Carpinus and *Larix* when assessed as a proportion of total range (Table 2), and range losses are high generally, with an average of `r round(mean(loss_table$Loss), 1)`% across all taxa, *Larix*, *Ulmus* and *Betula* have the largest absolute losses in range within the Upper Midwest.  Gains, conversely, account for no more than `r round(max(loss_table$Gain),1)`% of any taxon's historical range, with the highest gains observed fo *Populus* (Table 2) and averages `r round(mean(loss_table$Gain),1)`% across all taxa.   Spatial patterns of range gains vary widely among tree genera, and often show less spatial bias than losses (Figure 3).  Range gains for *Picea*, *Larix*, *Abies*, *Tsuga*, *Fagus*, *Ostrya*/*Carpinus*, *Tilia*, and *Fraxinus*, for example, all show a dispersed pattern of gains in a relatively few and widely scattered cells, suggesting that these apparent range gains are mainly caused by sampling uncertainty inherent in local-scale plots and stand heterogeneity, rather than systematic range gains.  Conversely, gains for *Populus* and *Acer*, are concentrated in northern Minnesota, Wisconsin, and Michigan, suggesting range infilling, while *Quercus* gains are consistent with some range extension northwards.  *Pinus*, *Thuja*, and *Ulmus* gains are limited but primarily in the southern portion of their historic range, or beyond it.

@@

Both the total change in realized climate distributions and the attribution of these changes to historic land use or climate change varies strongly among taxa and among climate variables (Figure 5). Shifts in realized climate distributions over the last 100 to 150 years are largest for *Ulmus* (TJan, TJul), *Larix* (Tmax), Tsuga (Tmax and Pann), and Fagus (Pann).   Attribution analyses based on the d index (Figure 5) reveal that while many genera show the imprint of historic land use on shifting climatic niches, this shift is not uniform across taxa. Maximum temperature shows the greatest land use signal (Figure 5X), with effects particularly strong for *Larix*, *Ulmus*, and Fagus. The large shift in the distribution of Tsuga relative to July temperatures, conversely, appears to be mainly attributable to historic climate change. For minimum temperature (Figure 5X), shifting distributions of *Ulmus* and *Thuja* can be mainly attributed to land use change, while *Larix* and *Betula* have large shifts that are equally attributable to historic climate change and land use, and the niche shifts for other taxa are driven by changing climate.  The climatic signal is strongest for annual precipitation, with regional changes in annual precipitation being largely responsible for shifting climate niches (Figure 5X), while only *Populus* and *Picea* have seen climate niche shifts due to land-use that are larger than the climate shifts.

@@

This work demonstrates that historic land use can substantially alter the realized climate niches of species and adds to our understanding of the interacting effects of climate change and land-use [@pyke2004habitat]. Land use change reduces the correlational structure between species and climate [@devineau2011arid] and is likely to have one of the strongest effects on changes in global biodiversity over the next century [@sala2000global]. While rates of land-use change may be declining in North America and in the upper Midwest [@dale1997landuse; @rhemtulla2009legacies] the legacies of land-use are likely to persist, both in terms of forest composition [@goring2015preset], and the correlational structure between species and climate. This work reinforces the importance of historical datasets for examining the relationships between species and climate, and their importance in forecasting species distributions in the future [@moritz2013future;@maguire2015modeling]. Modern surveys sample only a subset of historical distribution, both spatially and within multidimensional climate space.

Previous studies have shown that the realized climate niches of tree species also shifted during the climate changes accompanying the last deglaciation.  To a first order, many species accommodated past climate changes by shifting their ranges; i.e. they maintained a relatively stable distribution in environmental space by shifting their distributions within geographic space.  These past range shifts are well documented by paleodata and generally well simulated by species distribution and community-level models driven by paleoclimatic simulations [@martinez2006conservatism;@maguire2015modeling;@prentice1991vegetation].  However, species tracking of past climatic change was imperfect and other studies indicates shifts in species realized over time [@maiorano2013building;@pearman2008niche].  Mechanisms for these realized niche shifts include species lagging rapid climate change [@ordonez2013realized], species occupying non-analogue climates [@veloz2012no], or shifts in competition resulting from changing patterns of co-occurrence through time [@maiorano2013building].  This study adds to this prior work by showing that the large land use changes of the last several centuries can also significantly modify species-climate relationships.

In this study, shifts in realized climate niches and the underlying interactions between climate change and land use emerge from three basic processes.  First, climates have changed over the 20th and early 21st centuries, with rising winter minimum temperatures, declining summer maximum temperatures, and increasing precipitation.  Second, tree species distributions have not yet fully adjusted to these climate changes, causing the environmental niches to shift.  Third, anthropogenic land use has selectively occupied some portions of geographic and environmental space, making these regions less available to tree genera.  Because much of the land-use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009historical], the effects of climate change and land use are compounding.  Northern species show shifts away from drier habitat, while land-use in the northern prairie-forest margin intensifies, shifting species further than expected along the precipitation gradient. For southern tree species, land-use conversion has primarily occurred along the southern margins of the region, parallel to temperature and precipitation gradients.

Some of these apparent shifts in climate niches may be due to differences in sampling design rather than to actual shifts in species distributions.  In particular, the PLS survey design resulted in more survey points across the landscape than in the FIA dataset, but fewer trees surveyed per point.  The FIA is more confined to forested and semi-forested areas because plot surveys are limited to plots with > 10% tree cover [@woudenberg2010forest].  Hence, the FIA dataset may be underrepresenting tree distributions in unforested or semi-forested regions.  Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States [@wang2016landscape;@nieto2015close;@bell2014early].  Hence, any incompleteness in the representation of realized climate niches based on FIA data may result in inaccurate model projections.

On the other hand, anthropogenic agency may be extending the range of some tree species, for example, as a result of irrigation or landscaping care within urban or rural locations that might otherwise not support forest cover [@brazel2000tale].  For example, human agency causes urban ecosystems in deserts to closely resemble those in the northeastern United States [@groffman2014ecological].  Human agency may also support the extension of forest ecosystems at historical range margins, or preserve historical range margins from land use conversion to agriculture as windrows, or isolated woodlots. Detailed sampling of tree species, particularly in urban and agricultural environments beyond the current expected range limits of species, may provide better representation of species' fundamental climate niches or tolerance niches [@sax2013niche].  However, usage of species distributional data from highly managed ecosystems must account for the intensive management that may be required to maintain species beyond their realized niche.

These compounding and confounding interactions between historic climate change and land use, and the consequent shifts in realized climate space have important implications for the management and genetics of forest communities.  On the one hand, the observation that historical realized niches are broader than modern niches suggests that some tree genera may have more resilience to climate change.  Critically, the combination of regional land-use change and climate change based on contemporary data alone leads to the under-prediction of species resilience to climate change regionally. Regionally, annual temperature is expected to increase in Wisconsin by 3.6^o^C in 2055 from a 1980s baseline [@wicci2011climate]. Every tree genera examined in this study had historical $T_{max}$ values between 0.4 and 1.3^o^C higher than modern estimates, and with $P_{ann}$ between 42 and 86mm lower than modern estimates. Projected changes in $P_{ann}$ range from -17 - 110mm from a 1980 baseline to 2055 based on fourteen global climate models compiled as part of the Climate Model Intercomparison Project [@wicci2011climate]. Higher temperatures and lower precipitation means for taxa for the pre-settlement period indicate the potential for more resilience to climate change than might be assessed using modern data only.  This effect may be critical for species that reach their northern or western range limits in the upper Midwestern United States (e.g., Tsuga, Fagus), because a narrowing of the apparent environmental envelope, due to the twin action of land-use change and changing climate, may result in underprediction of species resilience to climate change.

However, projected climate change far exceeds the shifts due to historical land use and climate change.  In addition, coupled migration and adaptation in response to future climate change is likely to be impeded by the extirpation of communities due to land use change and loss of genetic diversity.  Species migration and genetic adaptation are coupled across time-scales [@davis2001range].  The loss of forest communities along the forest-prairie boundary in the upper Midwest is particularly important given that the direction of climate change regionally is towards warmer conditions, conditions that exists at the forest-prairie boundary.  The loss of genotypes adapted to warmer and drier conditions (the lagging edge of distributions for many taxa in the Upper Midwest) may mean more rapid losses at the lagging edges of the species range.  Long distance gene flow of adaptive traits to central and leading populations from the lagging edge [@hu2006seed] is lost when land use conversion to agricultural production results in the extirpation of locally adapted populations.  Thus net gene flow in populations along the remaining "lagging" edge will be from the central and leading populations, individuals maladapted to warmer conditions. This maladaptive gene flow could potentially amplify the effects of climate change on populations along the forest-prairie boundary leading to increasing risk of local extinction among natural populations [@kremer2012long].

Species distribution models fall along a continuum from more correlative to more process-based [@dormann2012correlation], and correlational species-distribution models are likely to be the most affected by historical land-use change and the sampling design of modern survey data. However, even process based models are partially parameterized from contemporary distributions, particularly with respect to the environmental tolerances of plant functional types [*e.g.*, @wullschleger2014plant].  Spatially biased land-use conversion could result in the extirpation of well-adapted genotypes within a population. In this case, both correlative and process based models might be affected by spatially biased land-use conversion.

In summary, historical datasets and multitemporal studies such as this oneadd to our ability to understand species realized climate niches and their adaptive potential in the face of climate change. They also serve to underscore the important interactions between climate change and land use change at a population level.  The strong changes in climate and land use over the past two centuries, combined with the likelihood that tree distributions are lagging recent climate changes [@svenning2013disequilibrium], create the strong likelihood that tree-climate distributions have shifted and that current tree-climate distributions are not representative of pre-settlement conditions.  Given the importance of modern distributional datasets for predicting species range shifts, compounding may result in the impression of narrower climatic tolerances than actually exist, leading to underestimates of species resilience.  At the same time, confounding interactions between historic climate change and land-use effects may increase predictive uncertainty.
